
# Exercise 6. Analysis of longitudinal data

*The end.*

**Exploring the data**

In the BPRS dataset 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated
on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations
and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia [(Vehkalahti & Everitt 2019)]("https://researchportal.helsinki.fi/en/publications/multivariate-analysis-for-the-behavioral-sciences-second-edition").

RATS dataset includes data from a nutrition study conducted in three groups of rats [(Crowder and Hand, 1990)]("https://doi.org/10.1201/9781315137421"). The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly over a 9-week period. 


```{r, echo=FALSE}

date()
```

```{r summary}
library(data.table)

BPRSL <- fread("F:/IODS_usb/IODS-project/data/BPRSL.csv")
RATSL <- fread("F:/IODS_usb/IODS-project/data/RATSL.csv")
BPRS <- fread("F:/IODS_usb/IODS-project/data/BPRS.csv")
RATS <- fread("F:/IODS_usb/IODS-project/data/RATS.csv")

str(BPRSL)
str(RATSL)
str(BPRS)
str(RATS)

# converting categorical variables of both dataset into factors
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
```

**Plotting the non-standardized weights of the diet groups of the rats **

On the RATS data diet group 1 had clearly lower weights in comparison to diet groups 2 and 3. In all groups it seemed that weights had increasing trend during the diet periods. In diet group 2 there was one individual with clearly higher weight than the others and diet groups 1 and 3 one individual each that had lower weight than the others. 

```{r ggplot61, fig.cap="Weights by diet group for the RATS data"}
#Access the package ggplot2
library(ggplot2)
library(tidyr)
library(dplyr)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))

```

**Plotting the standardized weights of rat diet groups**

After standardizing the weights of the groups there were no clear indication anymore of increasing trend of the weights. In group 1 it looks like the weights had stayed the same, in group 2 had decreasing and 1 increasing trend and in group 3 it looked like 3 individuals had decreasing trend and 1 (outlier) had increasing trend. Possible outliers in each groups were still detectable.
```{r ggplot62, fig.cap="Standardized weights of rat diet groups"}

# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised weight
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```

**Plotting the mean profiles of diet groups**
  
Mean weights of all groups were increasing during the diet. Group number 1 was most uniform having smallest standard deviation, although mean weight of the group was clearly the smallest (< 300 grams?). Group number 3 had highest average weight (> 500 grams) from start to end, but it looked like group number 2 mean weight was closing to group 3 average in the end. Standard deviation was the highest in group 2. 
  
```{r ggplot63, fig.cap="Mean response profiles for the three diet groups in the RATS data"}

# Number of time points, baseline (week 1) included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of weight by diet group and time 
RATSS <- RATSL %>%
  group_by(Group,Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```

**Finding outliers**

Looking at the boxplots created from each groupd there was indeed one outlier in each group

```{r ggplot64, fig.cap="Boxplots for the RATS diet groups during the WD 8-64"}

# Create a summary data by Group and ID with mean as the summary variable (ignoring WD 1).
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus diet group
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), WD 8-64")
```

**Removing outliers**

Outliers were removed after identifying their ID's

```{r, ggplot65, fig.cap="Boxplots for the RATS diet groups during the WD 8-64 after getting rid of the outliers"}

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- RATSL8S %>% filter(ID != 2 & ID != 12 & ID != 13)

# Draw a boxplot of the mean versus diet group
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), WD 8-64")
```

**Analysis of variance (ANOVA)**

ANOVA test showed that baseline mean weight of rats was statistically different (p < 0.01) from the mean weight of the WD 8-64. Groups 2 and 3 were significantly different (p-values < 0.01) from group 1, but there were no significant difference between groups 2 and 3 (p-value > 0.05.

```{r}

# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

# computing anova between diet groups
fit2 <- aov(mean~Group, data = RATSL8S2)
TukeyHSD(fit2)
```

**Plotting BPRSL data**

It looks like there was significant (p-value < 0.01) reduction in bprs values during the 9 week period but there was no significant difference between treatment group 1 and 2 (p-value > 0.05)


```{r ggplot66, fig.cap="Bprs values over 9 week period"}

# dim(BPRSL)

# create new variable to distinguish subjects of treatment groups

BPRSL$sID <- as.factor(paste(BPRSL$treatment, BPRSL$subject, sep = ""))

# plot the BPRSL data
ggplot(BPRSL, aes(x = week, y = bprs, group = sID)) +
  geom_line(aes(color = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)

```

**Random intercept model**

In the random intercept model, the linear regression fit intercept of each subject can vary from other subjects.  
There were no big difference in intercept and slope estimates in comparison to linear model, but standard errors of intercept and treatment 2 were higher and the week variable's lower in random intercept model.

```{r}

# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | sID), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

**Random intercept and random slope model**

In a random intercept and random slope model the linear regression fits for each subject can vary in intercept but also in slope from other subjects. This way it is possible to account for the individual differences in the subjects' bprs profiles, but also study the effect of time.

Treatment 2 slope estimate (1.51) was now much higher in comparison to previous linear and random intercept model (both 0.57). Random intercept and slope model was significantly different (p-value < 0.01) when compared to random intercept model.

```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | sID), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

**Random intercept and random slope model with interaction**

In our final model we fit a random intercept and slope model that allows for a group × time interaction. 
In this model slope estimate of treatment 2 is now negative (-2.3) and standard reached the highest value (4.2) in comparison to previous models. 

Interaction model is not significantly (p-value > 0.05) different from BPRS_ref1 model. In conclusion treatment does lead improved bprs values over time, but does not have interaction with time.  

```{r ggplot67, fig.cap="Fitted bprs values over 9 week period"}

# create a random intercept and random slope model
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | sID), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the four models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRSL
ggplot(BPRSL, aes(x = week, y = bprs, group = sID)) +
  geom_line(aes(color = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs values") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>%
  mutate(Fitted)

# draw the plot of the fitted BPRSL
ggplot(BPRSL, aes(x = week, y = Fitted, group = sID)) +
  geom_line(aes(color = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")

```


